Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2015-07-06 futurework.tex
Add some consistency macros

In this paper, we have only considered the parsing of strings, but dates are often formatted as integers (e.g. 201507016). It is significantly faster to decompose integers into date parts using arithmetic operations (e.g. mod and div) than by using slower, locale-sensitive string parsing functions. Moreover, the number of possible formats is low enough that we may be able to enumerate them. Timestamps are another form of numeric date for which we intend to automate the preparation.
In the course of our research, we have also identified a number of date part variants (e.g. ordinal dates, 4 letter month abbreviations, alternate meridian markers and postfix quarter syntax) that we would like to commend to the ICU project and maybe provide implementations for.

>>>>

In this paper, we have only considered the parsing of strings, but dates are often formatted as integers (\eg \texttt{201507016}). It is significantly faster to decompose integers into date parts using arithmetic operations (\eg \texttt{mod} and \texttt{div}) than by using slower, locale-sensitive string parsing functions. Moreover, the number of possible formats is low enough that we may be able to enumerate them. Timestamps are another form of numeric date for which we intend to automate the preparation.

In the course of our research, we have also identified a number of date part variants (\eg ordinal dates, four-letter month abbreviations, alternate meridian markers and postfix quarter syntax) that we would like to commend to the ICU project and maybe provide implementations for.

### 2015-07-06 intro.tex
Add some consistency macros

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as \cite{Polaris,Qlik,Spotfire}. While computers can provide high-speed and high-volume data processing, humans have domain knowledge and the ability to process data in parallel, notably by using our visual systems. Most importantly, humans provide the definition of what is valuable in an analysis. Accordingly, human/computer analytic systems are essential to extracting knowledge of value to humans and their societies from the large amounts of data being generated today.

\subsection{Interactivity}
Visualization systems are most effective when they are interactive, thereby allowing a user to explore data and connect it to their domain knowledge and sense of what is important without breaking cognitive flow. In recent years, a number of such systems have been developed, both by the academic community and by the commercial sector. Exploration of data consists not only in creating visual displays, but also in creating and modifying domain-specific computations. Consequently, most data visualization systems include facilities for defining such calculations as part of the data model being analyzed. The most effective systems allow users to define these calculations as part of the analytic interaction, which permits the user to stay in the flow of analysis \cite{Morton}.

During the analytic process, a user may discover that parts of the data are not yet suitable for analysis. Solutions to this problem are often provided by data preparation tools external to the visual analysis environment, which requires the user to break their analytic flow, launch another tool and reprocess their data before returning to their analysis. If the user does not own this process (\textit{e.g.} it is the responsibility of another department), then there can be significant delays (including ``never.'') More subtly, the result of updated external processing may not be compatible with the user's existing work, which can lead to more time lost reconciling the new data model with the existing analysis.

@@

\subsection{The Tableau Ecosystem}
In this paper, we shall be looking at an example of this problem in the context of the Tableau system. Tableau is a commercial visual analysis environment derived from the Polaris \cite{Polaris} system developed at Stanford. In addition to an interactive desktop application for creating data visualizations, the Tableau ecosystem also includes servers for publishing and sharing interactive visualizations. These servers can be privately operated by individual customers or hosted in the cloud.

\subsubsection{Tableau Public}
Of particular relevance in the present work is a free version of the server environment called Tableau Public \cite{Public}. Tableau Public (or ``Public'') allows authors to publish data visualizations that can be shared with the general public. The published data sets are limited to 100K rows and must be stored using the Tableau Data Engine (or ``TDE'')\cite{TDE1,TDE2}, but are otherwise free to use the full analytic capabilities of the Tableau system. In particular, the TDE provides native support for all analytic calculations defined by the authoring and publishing components. Moreover, visualizations published to Public are uploaded as complete Tableau workbooks, including the data model developed as part of the analysis. This makes it a rich source of data on the analytic habits and frustrations of Tableau users.

@@

\subsection{DATEPARSE}
Among the recent additions to the Tableau function library is a function called \texttt{DATEPARSE}, the usability of which is the focus of the work in this paper.

@@

The SQL-99 standard defines three temporal scalar types: \texttt{DATE}, \texttt{TIMESTAMP} and \texttt{TIME}. They are typically implemented as fixed-point types containing an offset from some epoch (\textit{e.g.} Julian Days.) This makes them compact to store and allows some temporal operations to be implemented very efficiently using arithmetic. Thus from the RDBMS perspective, representing dates in scalar form provides numerous benefits for users, both in terms of available operations and query performance.

Tableau models the first two of these types as ``Date'' and ``Date \& Time''; the third (pure time) is folded into Date \& Time by appending it to a fixed date of \texttt{1899-12-30}. From the analytic perspective, date types are dimensional (\textit{i.e.} independent variables) and can be used as either categorical (simply ordered) or quantitative (ordered with a distance metric) fields.

@@

\subsubsection{Parsing}
One of the oldest data preparation problems observed in Tableau is the parsing of date scalars so that users can perform these kinds of analyses. Training materials from the early days of the company included examples of how to convert columns of integers in the form \texttt{yyyyMMdd} to date scalars. The documented solution at the time was to convert the integer to a string and perform some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach had a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower)
\item Default parsing of date formats is locale-dependent, and may not work when the workbook is shared across an international organization (\textit{e.g.} between the US and European offices)
\item The expression code was hard to understand and maintain because it used a verbose, general-purpose string-handling syntax instead of a domain language.
\end{itemize}

@@

Our first solution to this problem was to add a new function to the Tableau calculation language called \texttt{DATEPARSE}. This function would take a string column and convert it to a datetime using a special purpose domain language for describing dates. Such functions exist in a number of databases (\textit{e.g.} Oracle, Postgres and MySQL) and adding it to the TDE was straightforward, so it was a natural addition to the function library.

We chose the date parsing syntax defined by the International Components for Unicode (or ICU) project \cite{icu} for the common domain language because it mirrored what was available in the Tableau code base, and translating this syntax into the formats used by various database vendors (\textit{e.g.} MySQL, Oracle, Postgres) was relatively painless. There are a few patches of non-overlapping functionality, but they tend to be obscure and we provide warnings when functionality is missing.

\subsubsection{	Usability}
After shipping this new function to our user base, we investigated how it was being applied in the field. We examined a few months of workbooks that had been uploaded to Public after the release to see if and how it was being used. We found that although there were a number of new calculations using \texttt{DATEPARSE}, the error rate for the domain language syntax was about 15\%. \texttt{DATEPARSE} was capable of solving the problem, and some users were able to discover it, but the syntax did not appear to be easy to use reliably.

@@

\subsection{Related Work}
Data preparation has been a known analytic bottleneck since at least the description of the Potter's Wheel system \cite{PottersWheel}. Since then, several other interactive data preparation systems have been proposed \cite{Wrangler,Refine}. While effective, these systems all make assumptions about possible date formats (\textit{e.g.} the domain system in \cite{PottersWheel}), which we suggest are too restrictive for real world data.

The Minimum Descriptive Length technique was first proposed in \cite{mdl}.  We have built on the version described in \cite{PottersWheel}.

>>>>

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as \cite{Stolte:2008,Qlik,Ahlberg:1996}. While computers can provide high-speed and high-volume data processing, humans have domain knowledge and the ability to process data in parallel, notably by using our visual systems. Most importantly, humans provide the definition of what is valuable in an analysis. Accordingly, human/computer analytic systems are essential to extracting knowledge of value to humans and their societies from the large amounts of data being generated today.

\subsection{Interactivity}
Visualization systems are most effective when they are interactive, thereby allowing a user to explore data and connect it to their domain knowledge and sense of what is important without breaking cognitive flow. In recent years, a number of such systems have been developed, both by the academic community and by the commercial sector. Exploration of data consists not only in creating visual displays, but also in creating and modifying domain-specific computations. Consequently, most data visualization systems include facilities for defining such calculations as part of the data model being analyzed. The most effective systems allow users to define these calculations as part of the analytic interaction, which permits the user to stay in the flow of analysis~\cite{Morton:2012}.

During the analytic process, a user may discover that parts of the data are not yet suitable for analysis. Solutions to this problem are often provided by data preparation tools external to the visual analysis environment, which requires the user to break their analytic flow, launch another tool and reprocess their data before returning to their analysis. If the user does not own this process (\eg it is the responsibility of another department), then there can be significant delays (including ``never.'') More subtly, the result of updated external processing may not be compatible with the user's existing work, which can lead to more time lost reconciling the new data model with the existing analysis.

@@

\subsection{The Tableau Ecosystem}
In this paper, we shall be looking at an example of this problem in the context of the Tableau system. Tableau is a commercial visual analysis environment derived from the Polaris~\cite{Stolte:2008} system developed at Stanford. In addition to an interactive desktop application for creating data visualizations, the Tableau ecosystem also includes servers for publishing and sharing interactive visualizations. These servers can be privately operated by individual customers or hosted in the cloud.

\subsubsection{Tableau Public}
Of particular relevance in the present work is a free version of the server environment called Tableau Public. Tableau Public (or ``Public'') allows authors to publish data visualizations that can be shared with the general public. The published data sets are limited to 100K rows and must be stored using the Tableau Data Engine (or ``TDE'')~\cite{Wesley:2011,Wesley:2014}, but are otherwise free to use the full analytic capabilities of the Tableau system. In particular, the TDE provides native support for all analytic calculations defined by the authoring and publishing components. Moreover, visualizations published to Public are uploaded as complete Tableau workbooks, including the data model developed as part of the analysis. This makes it a rich source of data on the analytic habits and frustrations of Tableau users.

@@

\subsection{DATEPARSE}
Among the recent additions to the Tableau function library is a function called \dateparse, the usability of which is the focus of the work in this paper.

@@

The SQL-99 standard defines three temporal scalar types: \texttt{DATE}, \texttt{TIMESTAMP} and \texttt{TIME}. They are typically implemented as fixed-point types containing an offset from some epoch (\eg Julian Days.) This makes them compact to store and allows some temporal operations to be implemented very efficiently using arithmetic. Thus from the RDBMS perspective, representing dates in scalar form provides numerous benefits for users, both in terms of available operations and query performance.

Tableau models the first two of these types as ``Date'' and ``Date \& Time''; the third (pure time) is folded into Date \& Time by appending it to a fixed date of \texttt{1899-12-30}. From the analytic perspective, date types are dimensional (\ie independent variables) and can be used as either categorical (simply ordered) or quantitative (ordered with a distance metric) fields.

@@

\subsubsection{Parsing}
One of the oldest data preparation problems observed in Tableau is the parsing of date scalars so that users can perform these kinds of analyses. Training materials from the early days of the company included examples of how to convert columns of integers in the form \texttt{yyyyMMdd} to date scalars. The documented solution at the time was to convert the integer to a string and perform some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach had a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower)
\item Default parsing of date formats is locale-dependent, and may not work when the workbook is shared across an international organization (\eg between the US and European offices)
\item The expression code was hard to understand and maintain because it used a verbose, general-purpose string-handling syntax instead of a domain language.
\end{itemize}

@@

Our first solution to this problem was to add a new function to the Tableau calculation language called \dateparse. This function would take a string column and convert it to a datetime using a special purpose domain language for describing dates. Such functions exist in a number of databases (\eg Oracle, Postgres and MySQL) and adding it to the TDE was straightforward, so it was a natural addition to the function library.

We chose the date parsing syntax defined by the International Components for Unicode (or ICU) project~\cite{ICU} for the common domain language because it mirrored what was available in the Tableau code base, and translating this syntax into the formats used by various database vendors (\eg MySQL, Oracle, Postgres) was relatively painless. There are a few patches of non-overlapping functionality, but they tend to be obscure and we provide warnings when functionality is missing.

\subsubsection{	Usability}
After shipping this new function to our user base, we investigated how it was being applied in the field. We examined a few months of workbooks that had been uploaded to Public after the release to see if and how it was being used. We found that although there were a number of new calculations using \dateparse, the error rate for the domain language syntax was about 15\%. \dateparse was capable of solving the problem, and some users were able to discover it, but the syntax did not appear to be easy to use reliably.

@@

\subsection{Related Work}
Data preparation has been a known analytic bottleneck since at least the description of the Potter's Wheel system~\cite{Ramen:2001}. Since then, several other interactive data preparation systems have been proposed, including Data Wrangler~\cite{Kandel:2001} and Google refine~\cite{Refine}. While effective, these systems all make assumptions about possible date formats (\eg the domain system in~\cite{Ramen:2001}), which we suggest are too restrictive for real world data.

The Minimum Descriptive Length technique was first proposed in Rissanen~\cite{Rissanen:1978}.  We have built on the version described in~\cite{Ramen:2001}.

### 2015-07-06 paper.tex
More formatting
Redo all the quotes (I’d forgotten how annoying TeX’s quote mark
conventions are). Format e.g. everywhere (next time I’ll define a
macro…)
Richard Bibliography
Add some consistency macros

%
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%

\begin{document}

@@

\title{Does Anybody Really Know What Time It Is?\\
Automating the Extraction of Date Scalars}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

One of the most common parsing tasks is extracting date and time data from string representations. Several databases include date parsing ``mini-languages" to cover the wide range of possible formats. Tableau has recently provided a function called \texttt{DATEPARSE}  with a single syntax that is translated into the nearest equivalent in the underlying database. Subsequent analysis of customer usage of this function shows that the parsing language syntax was difficult for users to master.

@@

\keywords{Date parsing, automated cleaning, structure extraction} % NOT required for Proceedings

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
We would like to thank to Pat Hanrahan for encouraging this work; Roger Hau for providing the resources to productize it; and Jason King, whose harrowing live demonstration of manual \texttt{DATEPARSE} in front of several thousand people inspired this project.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{dateparse}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%

>>>>

%
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%
\def\eg{\textit{e.g.} }
\def\ie{\textit{i.e.} }
\def\dateparse{\texttt{DATEPARSE} }
%
\begin{document}

@@

\title{Does Anybody Really Know What Time It Is?\\
Automating the Extraction of Date Scalars}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% ``name/affiliations'' may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

One of the most common parsing tasks is extracting date and time data from string representations. Several databases include date parsing ``mini-languages'' to cover the wide range of possible formats. Tableau has recently provided a function called \dateparse with a single syntax that is translated into the nearest equivalent in the underlying database. Subsequent analysis of customer usage of this function shows that the parsing language syntax was difficult for users to master.

@@

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
We would like to thank to Pat Hanrahan for encouraging this work; Roger Hau for providing the resources to productize it; and Jason King, whose harrowing live demonstration of manual \dateparse in front of several thousand people inspired this project.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{dateparse}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ``.bib'' file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%

### 2015-07-06 parameters.tex
Add some consistency macros

\subsubsection{NULL Filtering}
Each column was then reduced to a maximum sample set of 32. We excluded domain values that appeared to be representations of \texttt{NULL}:
\begin{itemize}
\item Values containing the substring \texttt{NULL}
\item Values containing no digits and at most one non-whitespace character (e.g. \texttt{" / / "})
\item Values on a list of empirically determined common \texttt{NULL} values (e.g. \texttt{0000-00-00}, \texttt{NaN})
\end{itemize}
Columns that had no remaining valid samples were discarded.

@@

The syntax is documented at the ICU web site \cite{icu}. While fairly complete, it has a few limitations that we ran into when evaluating the algorithms:
\begin{itemize}
\item No support for 4 letter year abbreviations (e.g. \texttt{Sept.})
\item No support for ordinal days (e.g. \texttt{July 4th})
\item No support for quarter postfix notation (e.g. \texttt{2Q})
\item No support for variant meridian markers (e.g. \texttt{a.m.})
\end{itemize}

@@

One other quirk of the ICU syntax may be a contributing factor to the user confusion around writing correct ICU date formats. The use of lexicographical case in the meta-symbols of the format language can be confusing (e.g. \texttt{y} is used for years, but \texttt{M} is used for months while \texttt{m} is used for minutes.) Another advantage of an automated algorithm is that it hides such problems from most users, significantly improving the usability of the function.

>>>>

\subsubsection{NULL Filtering}
Each column was then reduced to a maximum sample set of 32. We excluded domain values that appeared to be representations of \texttt{NULL}:
\begin{itemize}
\item Values containing the substring \texttt{NULL}
\item Values containing no digits and at most one non-whitespace character (\eg \texttt{" / / "})
\item Values on a list of empirically determined common \texttt{NULL} values (\eg \texttt{0000-00-00}, \texttt{NaN})
\end{itemize}
Columns that had no remaining valid samples were discarded.

@@

The syntax is documented at the ICU web site~\cite{ICU}. While fairly complete, it has a few limitations that we ran into when evaluating the algorithms:
\begin{itemize}
\item No support for 4 letter year abbreviations (\eg \texttt{Sept.})
\item No support for ordinal days (\eg \texttt{July 4th})
\item No support for quarter postfix notation (\eg \texttt{2Q})
\item No support for variant meridian markers (\eg \texttt{a.m.})
\end{itemize}

@@

One other quirk of the ICU syntax may be a contributing factor to the user confusion around writing correct ICU date formats. The use of lexicographical case in the meta-symbols of the format language can be confusing (\eg \texttt{y} is used for years, but \texttt{M} is used for months while \texttt{m} is used for minutes.) Another advantage of an automated algorithm is that it hides such problems from most users, significantly improving the usability of the function.

### 2015-07-06 testing.tex
Add some consistency macros

\begin{table}[ht]
\begin{tabular}{|p{0.48\linewidth}| p{0.48\linewidth}|}
\hline
\textbf{Number of Records} & 31,546\\ \hline
\textbf{Error Rate} & 27.95\% \\ \hline
\textbf{Analysis Speed ($\mu s$)} & 2,245.04 \\ \hline
\textbf{Validation Speed ($\mu s$)} & 1.65 \\ \hline
\textbf{Median Not Null} & 50 \\ \hline
\end{tabular}
\label{tab:mdlstats}
\caption{MDL Parsing Statistics.}
\end{table}

@@

Some of the dates are clearly just numbers, but our approach is to assume that when the user tells us that the column contains dates we should find the best fit. The samples include dates from a wide range of historical sources (e.g. Roman pottery dates) so we have elected to defer the date identification task to the user.

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{|p{0.48\linewidth}| p{0.24\linewidth}|}
\hline
\textbf{Number of Records} & 31,546\\ \hline
\textbf{Error Rate} & 27.95\% \\ \hline
\textbf{Analysis Speed ($\mu s$)} & 2,245.04 \\ \hline
\textbf{Validation Speed ($\mu s$)} & 1.65 \\ \hline
\textbf{Median Not Null} & 50 \\ \hline
\end{tabular}
\label{tab:mdlstats}
\caption{MDL Parsing Statistics.}
\end{table}

@@

Some of the dates are clearly just numbers, but our approach is to assume that when the user tells us that the column contains dates we should find the best fit. The samples include dates from a wide range of historical sources (\eg Roman pottery dates) so we have elected to defer the date identification task to the user.
